ANCOVA Comparison Simulations: Discrete Outcomes

### 2017-09-13 simulations/discrete_outcomes.Rmd
Add discrete outcome simulations and results

---
title: "ANCOVA Comparison Simulations: Continuous Outcomes"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Continuous data

Step 1:
We generated continuous potential outcomes.
We used two scenarios.
In the first, the treatment effect was constant across strata.
We drew a latent random variate $v_{ij}$ from the uniform distribution on $[-4, 4]$.
In the second scenario, the treatment effect varied across strata, and we drew the latent random variable according to
\begin{displaymath}
   v_{ij} \sim \left\{
     \begin{array}{lr}
       \text{Unif}[-4, -1] & : j=1\\
       \text{Unif}[-1, 1] & : j=2\\
       \text{Unif}[1, 4] & : j=3\\
     \end{array}
   \right.
\end{displaymath}

We generated independent and identically distributed errors $\varepsilon_{ij}$ and $\delta_{ij}$ and varied the error distribution. 
The errors were either standard normal (to mimic the usual ANCOVA assumptions),
$t$ distributed with two degrees of freedom,
standard lognormal,
or exponentially distributed, with scale parameter $1$ and shifted to have mean zero.
The observed $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$ were independent across $i$ and $j$.

The baseline value for individual $i,j$ was 
$$ X_{ij} = \frac{\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \varepsilon_{ij}$$.

Then we generated potential outcomes as

$$Y_{ij}(Z_{ij}) = \frac{(2Z_{ij} - 1)\gamma e^{v_{ij}} + e^{v_{ij}/2}}{2} + \delta_{ij}.$$

The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.

Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

@@

Step 3:
We repeat step 1 and 2 for different distributions on $\varepsilon$ and $\delta$.

@@

```{r generate_sample}
gen_y1 <- function(gamma, v, error, Z){
  y1 <- 0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error
  return(y1)
}

gen_y0 <- function(gamma, v, error){
  y0 <- 0.5*(-1*gamma*exp(v) + exp(v/2)) + error
  return(y0)
}

@@

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_y0(gamma, v, epsilon)
  Y1 <- gen_y1(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

# Gaussian error terms, constant additive treatment effect

```{r vary_effect_size_plot, fig.height=4}
load("continuous_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the effect size was zero, all five tests had the correct level of 5\%.
Power for all tests increased with the effect size, though the differenced permutation test lagged behind the rest.
This is unsurprising, as the correlation between baseline $X$ and outcome $Y$ was low:
it was highest at $0.33$ when $\gamma=0$ and decreased to approximately $0$ when $\gamma=0.2$.
The parametric and permutation linear model tests perform comparably and are plotted on top of each other, while the stratified permutation test has slightly more power.

# Gaussian error terms, heterogeneous treatment effect
```{r vary_effect_size_het_plot, fig.height=4}
load("continuous_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
Four of the five tests had the correct level of 5\%, while the stratified permutation test rejected only 1\% of the time.
Both the stratified and differenced permutation tests had lower power than the linear model tests as the effect size increased.
The correlation between baseline $X$ and outcome $Y$ was low but nontrivial (between 0.04 and 0.44), so this result makes some sense:
it is beneficial for power to control for the baseline in the linear model, but the correlation is too low to make the differencing method useful.

# Comparison of error distributions

We varied the distributions of the errors $\varepsilon$ and $\delta$ and examined the power of the five tests.
Recall that these are not errors in the linear model framework, but more like disturbances on the potential outcomes and baseline measures for each individual.
These disturbances affect the distributions of $X$ and $Y$, which may impact the power of the parametric ANCOVA, and affect the variances of $X$ and $Y$, which may make one method of controlling for the baseline more effective than another.
We fixed $\gamma=0.2$ and let the errors have a standard normal distribution, a $t$ distribution with 2 degrees of freedom, a standard log normal distribution, or an exponential distribution with parameter $1$, shifted to have mean zero.

Table 1 shows the empirical power (rejection rate among 10,000 simulations) for each of these error distributions and tests, where the $v_{ij}$ came from a single distribution across strata.
In all cases, the correlation between baseline and outcome was lower than 0.05 in magnitude.
Thus, the stratified permutation test, which omits the baseline measurement, had the highest power for all error distributions.
All three linear model based tests had roughly the same rejection rate.
The differenced permutation test had substantially less power than the other methods, due to the low correlation between baseline and outcome.
Power was highest for normally distributed errors and was lowest for $t$ distributed errors.

Table 2 shows the same results, but where the distribution of $v_{ij}$ varied across strata.
In this case, the three linear model based tests had the highest power for each error distribution.
While the correlation between baseline and outcome was still nearly zero, controlling for it in the linear model tended to increase precision.
The stratified permutation test lost power for this reason.
The differenced permutation test continued to have low power.
Interestingly, when the errors were $t$ or log normally distributed, the power of each test did not change much whether the treatment effects were constant or heterogeneous across strata.
However, power was lower for normally distributed errors when treatment effects varied across strata.
In this scenario, the linear model does not fully describe the relationship between baseline and outcome: 
interaction terms between baseline and stratum ID are needed in the model to capture this variation across strata.

```{r gaussian_homogeneous}
gaussian_homogeneous <- gamma_power[[which(abs(gamma_vec - 0.2) < 1e-6)]] 
```
```{r gaussian_heterogeneous}
load("continuous_outcomes/gaussian_heterogeneous.Rda") 
```
```{r t_homogeneous}
load("continuous_outcomes/t_homogeneous.Rda")
```
```{r t_heterogeneous}
load("continuous_outcomes/t_heterogeneous.Rda")

```
```{r ln_homogeneous}
load("continuous_outcomes/lognormal_homogeneous.Rda")

```
```{r ln_heterogeneous}
load("continuous_outcomes/lognormal_heterogeneous.Rda")
```
```{r exponential_homogeneous}
load("continuous_outcomes/exponential_homogeneous.Rda")
```
```{r exponential_heterogeneous}
load("continuous_outcomes/exponential_heterogeneous.Rda")

```{r power_summary, results = "asis"}
powers <- list(gaussian_homogeneous,
               t_homogeneous,
               lognormal_homogeneous,
               exponential_homogeneous,
               gaussian_heterogeneous,
               t_heterogeneous,
               lognormal_heterogeneous,
               exponential_heterogeneous)
summary05 <- t(sapply(powers, function(x) x[5,]))
summary05 <- data.frame(rep(c("Normal", "t(2)", "Log Normal", "Exponential"), 2), summary05)
colnames(summary05) <-  c("Errors", "ANCOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane") 
```

```{r power_summary_const_tab, results = "asis"}
summarytab_constant <- xtable(summary05[1:4,], digits = 3,
       caption = "Empirical power at level $0.05$ for simulated data with constant additive treatment effects",
       label = "tab:power_grid1") 
print(summarytab_constant, include.rownames=FALSE) 
```

```{r power_summary_het_tab, results = "asis"}
summarytab_het <- xtable(summary05[5:8, ], digits = 3,
       caption = "Empirical power at level $0.05$ for simulated data with heterogeneous treatment effects",
       label = "tab:power_grid2") 
print(summarytab_het, include.rownames=FALSE) 
```

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 0.9) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
    ylim(0, 0.9) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")

@@

p3
ggsave("../ms/fig/continuous_outcomes_simulation_power.pdf", p3, device="pdf", width=8)

>>>>

---
title: "ANCOVA Comparison Simulations: Discrete Outcomes"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

# Discrete data

Step 1:
We generated the covariate sand potential outcomes according to the methods in the continuous outcome section but discretized them by taking the floor.
Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

@@

@@

```{r generate_sample}
gen_y <- function(gamma, v, error, Z){
  y <- floor(0.5*((2*Z-1)*gamma*exp(v) + exp(v/2)) + error)
  return(y)
}

gen_x <- function(gamma, v, error){
  x <- floor(0.5*(-1*gamma*exp(v) + exp(v/2)) + error)
  return(x)
}

@@

  # Generate covariates
  Z <- rep(0:1, length.out = N)
  Y0 <- gen_x(gamma, v, epsilon)
  Y1 <- gen_y(gamma, v, delta, Z)
  return(data.frame(Y1, Y0, Z, v, stratumID, epsilon, delta))
}

@@

# Constant additive treatment effect

```{r vary_effect_size_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power <- lapply(gamma_res, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") 
```

First, we let the $v_{ij}$ have the same distribution across strata and generated normally distributed errors.
We varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
In the random populations that were generated, this corresponded to population average treatment effects of `r round(tr_effect, 2)[1:4]`, and `r round(tr_effect, 2)[5]`, respectively.
Figure 1 shows the empirical power (rate of rejection in the 10,000 simulations) at level 5\% for these increasing effect sizes.
When the effect size was zero, all five tests had the correct level of 5\%, though the stratified and differenced permutation tests rejected 3.5\% of the time.
The differenced permutation test had substantially lower power than the rest,
again because the correlation between baseline $X$ and outcome $Y$ was low.
In this case, the unadjusted stratified test had comparable power to the linear model based tests.

# Heterogeneous treatment effect

```{r vary_effect_size_het_plot, fig.height=4}
load("discrete_outcomes/vary_gamma_het_results.Rda")
gamma_vec <- seq(0, 0.2, by=0.05)
gamma_power_het <- lapply(gamma_res_het, function(x) apply(x, 2, compute_power))
plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") 
```
Next, we varied the distribution of $v_{ij}$ across strata and generated normally distributed errors.
Once again, we varied $\gamma$ from $0$ to $0.2$ in steps of $0.05$.
The corresponding population average treatment effects were of `r round(tr_effect_het, 2)[1:4]`, and `r round(tr_effect_het, 2)[5]`, respectively.
Figure 2 shows the empirical power at level 5\% for these increasing effect sizes.
The pattern here is nearly identical to Figure 2 in the continuous outcome simulations, so we leave out further comments.

```{r plot_for_ms, include=FALSE}
p1 <- plot_power_gamma(gamma_power, tr_effect, 0.05, "Power at level 5%, Constant Treatment Effect") + 
  ylim(0, 0.95 ) +
  theme(legend.position="bottom") + 
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  report_theme +
  ggtitle("Constant Treatment \n Effect")

p2 <- plot_power_gamma(gamma_power_het, tr_effect_het, 0.05, "Power at level 5%, Heterogeneous Treatment Effect") +
    ylim(0, 0.95) +
    report_theme +
    ggtitle("Heterogeneous Treatment \n Effect")

@@

p3
ggsave("../ms/fig/discrete_outcomes_simulation_power.pdf", p3, device="pdf", width=8)
